\section{Introduction}
The rapid evolution of automotive architectures has driven a transition from traditional Controller Area Network (CAN) buses to high-speed Automotive Ethernet, enabling bandwidth-intensive applications such as surround-view camera systems, LiDAR data transmission, and over-the-air (OTA) software updates~\cite{matheus2021automotive}. While this architectural shift brings substantial functional advantages, it simultaneously expands the attack surface of in-vehicle networks by exposing vehicles to a broader class of cyber threats previously associated with enterprise and industrial networks, including port scanning, address spoofing, and payload injection~\cite{checkoway2011comprehensive}.

Intrusion detection systems (IDS) have emerged as a primary defense mechanism for in-vehicle networks, and a substantial body of research has been devoted to CAN-based IDS leveraging statistical anomaly detection, machine learning, and deep learning approaches~\cite{lokman2019intrusion}. However, Automotive Ethernet presents fundamentally different traffic characteristics: packets carry variable-length raw byte payloads, traffic is heterogeneous across network segments (e.g., control traffic versus high-bandwidth RTP camera streams), and the protocol stack closely mirrors standard IP networking. These properties render CAN-specific detection methods inapplicable and necessitate purpose-built approaches for Automotive Ethernet.

A central challenge in developing IDS for automotive systems is the scarcity of labeled attack data. Collecting representative attack traffic in real vehicles is operationally constrained, and manually annotating large-scale packet captures is prohibitively expensive. As a result, supervised learning approaches -- which have achieved strong performance in controlled benchmark settings -- are difficult to deploy in practice, as they require comprehensive coverage of attack types at training time and fail to generalize to unseen attack variants~\cite{sommer2010outside}. This motivates an unsupervised formulation of the intrusion detection problem, in which a model is trained exclusively on normal traffic and anomalies are identified as deviations from the learned normal behavior.

Self-supervised learning (SSL) has recently demonstrated remarkable success in learning rich representations from unlabeled data across vision~\cite{he2022masked}, language~\cite{devlin2019bert}, and time-series domains~\cite{yue2022ts2vec}. In particular, masked autoencoding -- wherein a model is trained to reconstruct randomly masked portions of its input -- has proven effective at capturing the underlying generative structure of complex sequential data. This property is well-suited to network traffic anomaly detection: a model that faithfully reconstructs normal packet sequences will produce elevated reconstruction errors when presented with anomalous traffic that deviates from learned patterns.

However, applying masked autoencoding to Automotive Ethernet traffic is non-trivial. Network packets contain two complementary sources of information that exhibit distinct statistical properties: the raw byte content of each packet, which encodes payload structure and protocol fields, and the inter-packet byte differences across a traffic sequence, which capture how each byte position evolves over consecutive packets. Modeling these two information sources with a single homogeneous encoder risks underutilizing either the global sequential structure or the per-byte temporal variation patterns. Furthermore, Automotive Ethernet carries heterogeneous traffic types within the same network, including low-bandwidth control packets and high-bandwidth RTP camera streams, each exhibiting distinct normal behavior and susceptibility to different attack classes.

To address these challenges, we propose a self-supervised dual-stream masked autoencoder for unsupervised intrusion detection in Automotive Ethernet networks. The proposed architecture processes each sliding window of 64 consecutive packets through two specialized encoding streams. The first stream applies a Transformer encoder to the normalized raw byte content of each packet, capturing global inter-packet dependencies across the full window; rather than the conventional fixed sinusoidal scheme, the Transformer's positional encoding is conditioned on the inter-packet interval recorded at capture time, allowing the model to jointly reason about packet content and arrival timing without any protocol-specific parsing. The second stream applies a residual 1D-CNN to the inter-packet byte difference sequence, where each byte position is treated as a channel, enabling the model to learn per-byte temporal dynamics across consecutive packets. The two streams are integrated through a delta-conditioned gated residual mechanism, in which gating weights derived from the CNN stream modulate its contribution to the Transformer representation, producing a unified latent vector through attention pooling. A conditioned sequence decoder then reconstructs the original packet content from this latent representation (Fig.~\ref{fig:architecture}). Training is performed exclusively on normal traffic using a packet-level masked reconstruction objective with asymmetric loss weighting that emphasizes the reconstruction of masked positions. At inference time, anomalies are flagged when the mean reconstruction error of a packet window exceeds a threshold derived from the normal traffic error distribution.

We evaluate the proposed method on the CarDS Automotive Ethernet dataset~\cite{hellemans2025cards}, which to our knowledge represents the first publicly available real-vehicle Ethernet capture dataset encompassing both control network and RTP camera stream traffic under multiple driving scenarios. The dataset includes a diverse set of attack scenarios comprising network scanning (Nmap SYN/ACK/FIN scans), IP and MAC address spoofing, and RTP stream replacement attacks, providing a comprehensive testbed for evaluating Automotive Ethernet IDS. Our method achieves an overall F1-score of 0.9802 (AUROC 0.9976, AUPRC 0.9974) across 27 attack files covering 23 attack types, with near-perfect detection of network scanning and spoofing attacks (F1 up to 0.9993) and competitive detection of RTP stream replacement attacks (F1 up to 0.9113). Ablation studies confirm the contribution of each architectural component, demonstrating that the dual-stream design consistently outperforms single-stream alternatives and that conditioning the positional encoding on real inter-packet timing further improves detection while substantially widening the range of operating thresholds over which performance remains stable.

The main contributions of this paper are as follows:
\begin{itemize}

\item To our knowledge, we present the first anomaly detection study conducted on the CarDS Automotive Ethernet dataset, establishing evaluation protocols and performance baselines for this newly available real-vehicle benchmark.

\item We propose a dual-stream masked autoencoder that jointly models global inter-packet sequential structure via a Transformer encoder and per-byte temporal dynamics via a residual CNN encoder applied to inter-packet byte differences, unified through a delta-conditioned gated residual fusion mechanism, enabling unsupervised intrusion detection without any labeled attack data.

\item We introduce a time-aware positional encoding that conditions the Transformer stream on real inter-packet capture timing rather than fixed sequence rank, requiring no protocol-specific parsing; this improves detection of timing-irregular attacks and substantially widens the range of detection thresholds over which performance remains stable, without materially increasing model size.

\item We design a packet-level masked reconstruction objective with asymmetric loss weighting that forces the model to learn the generative structure of normal Automotive Ethernet traffic from unlabeled data alone.

\item We conduct extensive experiments and ablation studies across 27 real-vehicle attack files spanning 23 attack types, demonstrating the effectiveness of the proposed architecture and providing detailed analysis of detection performance across heterogeneous attack categories.

\end{itemize}

The remainder of this paper is organized as follows. Section 2 reviews related work on in-vehicle network intrusion detection and self-supervised learning for anomaly detection. Section 3 describes the proposed method in detail. Section 4 presents the experimental setup, results, and ablation studies. Section 5 discusses the broader implications of these results, including the comparison with baseline methods, the role of RTP-specific behavior, and practical deployment considerations. Section 6 concludes the paper and outlines directions for future work.