\section{Discussion}
\label{sec:discussion}

\subsection{On the Baseline Comparison}
\label{sec:baseline_discussion}
% Table~\ref{tab:baseline} (Section~\ref{sec:baseline})는 완전 지도 학습~\cite{jeong2021convolutional}부터 준지도 학습~\cite{shibly2023feature}, 비지도 학습~\cite{jeong2023aero}에 이르기까지 레이블 요구 조건의 전체 스펙트럼을 아우르는 기존의 차량용 이더넷 IDS 접근 방식 3가지와 제안 기법을 비교하여 보여줍니다. 이 중 AERO~\cite{jeong2023aero}는 공격 레이블 없이 학습된다는 점에서 레이블 효율성 측면에서 가장 직접적으로 비교 가능한 모델입니다. 따라서 의미 있는 차별점은 비지도 학습 그 자체라기보다는, \emph{비지도 모델이 어떤 데이터를 바탕으로 학습하는가}에 있습니다. AERO의 특징 추출기는 학습 전에 프로토콜 지식이 개입된 3가지 수작업 특징(hand-crafted features)을 구성하는 반면, 제안된 방법은 원시 패킷 바이트에서 직접 작동합니다. 이러한 차이는 CarDS와 같이 다중 도메인 및 다중 프로토콜이 혼재된 환경에서 중요한 의미를 가집니다. 특정 프로토콜의 시맨틱(예: AVTP 또는 SOME/IP)에 맞춰진 특징 추출기는 다른 프로토콜이나 네트워크 세그먼트에 적용할 때 재설계가 필요할 수 있지만, 제안된 원시 바이트 표현은 구조적 설계 자체로 인해 세그먼트 전반에 걸쳐 범용적으로 일반화되기 때문입니다.
Table~\ref{tab:baseline} (Section~\ref{sec:baseline}) situates the proposed method relative to three existing Automotive Ethernet IDS approaches spanning the full spectrum of label requirements, from fully supervised~\cite{jeong2021convolutional} to semi-supervised~\cite{shibly2023feature} to unsupervised~\cite{jeong2023aero}. Among these, AERO~\cite{jeong2023aero} is the most directly comparable in terms of label efficiency, as it is likewise trained without attack labels; the meaningful distinction is therefore not unsupervised learning per se, but \emph{what} the unsupervised model is trained on. AERO's feature extractor constructs three hand-crafted, protocol-informed features prior to learning, whereas the proposed method operates directly on raw packet bytes. This distinction matters in a multi-domain, multi-protocol setting such as CarDS: a feature extractor tailored to one protocol's semantics (e.g., AVTP or SOME/IP) may require redesign when applied to a different protocol or network segment, whereas the proposed raw-byte representation generalizes across segments by construction. Quantitatively, this label-efficiency advantage manifests as a per-category complementarity in the baselines' failure modes: AERO, the only other label-free method, drops to near-random on Emergency/VLAN\,4 Nmap scans (AUROC 0.50) despite handling RTP replacement reasonably (0.89), whereas the semi-supervised classifier inverts this profile, remaining near-perfect on scans (AUROC $\approx$1.0) but weakest on RTP replacement (0.74--0.84). The proposed method attains strong detection on both categories (RTP 0.976, Nmap 0.9998) without any attack labels or protocol-specific features, which we attribute to a single raw-byte representation that generalizes across heterogeneous attack types where hand-crafted or partially labeled features do not.

\subsection{Analysis of RTP Stream Replacement Detection}
% 두 공격 카테고리 중, RTP 스트림 교체 공격은 RTP 헤더, 패킷 도착 간격, 페이로드 길이를 포함한 패킷 레벨 구조를 그대로 유지하면서 비디오 페이로드 콘텐츠만을 변조하기 때문에 구조적으로 고유한 탐지 난제를 제시합니다. 이로 인해 정상 트래픽과 이상 트래픽 간의 재구성 오차 마진이 헤더 레벨 공격보다 본질적으로 좁아, Section~\ref{sec:main_results}에서 보고한 바와 같이 임계값 독립 분리도(AUPRC 0.9044, Nmap 0.9998 대비)도 낮게 나타납니다. 카테고리 내의 파일별 F1-score 편차(0.8372 ~ 0.9113)는 탐지 가능성이 공격 메커니즘만이 아니라, 주입된 비디오 콘텐츠와 정상 카메라 트래픽 학습 분포 간의 통계적 거리에 의해서도 부분적으로 결정됨을 시사합니다. 학습 분포에 더 가까운 콘텐츠로 교체된 공격 파일은 탐지하기 더 어렵습니다. 이는 재구성 기반 이상 탐지에 내재된 특성이며, 공격자의 페이로드 선택에 의해 회피 난이도가 달라질 수 있음을 의미합니다.
Of the two attack categories evaluated, RTP stream replacement presents a structurally distinct detection challenge because it preserves packet-level structure --- valid RTP headers, consistent inter-arrival timing, and comparable payload lengths --- while substituting only the video payload content. This narrows the reconstruction error margin between normal and anomalous traffic, which is reflected in the threshold-independent separability reported in Section~\ref{sec:main_results} (AUPRC 0.9044 vs.\ 0.9998 for Nmap-based attacks). The per-file variation in F1-score within the category (0.8372--0.9113) further suggests that detectability is partly governed by the statistical distance between the injected video content and the training distribution of normal camera traffic, independent of the attack mechanism itself. Attack files whose replacement content is statistically closer to the normal distribution present a harder detection target --- an inherent property of reconstruction-based anomaly detection that implies evasion difficulty varies with the attacker's choice of replacement payload.

% 한 가지 흥미로운 관찰은, RTP 스트림 교체가 패킷 간 타이밍을 의도적으로 \emph{보존}함에도 불구하고 시간 인지형 위치 인코딩(Section~\ref{sec:pe_ablation})이 RTP 탐지 AUPRC를 $+0.043$ 향상시켰다는 점입니다. 이는 타이밍 기반 신호가 애초에 공격에 의해 변조되지 않기 때문에, 개선이 타이밍 이상 직접 탐지에서 비롯된 것이 아님을 시사합니다. 두 가지 비상호배타적 가설을 제시합니다. 첫째, 실제 IPI 데이터로 셀프 어텐션을 조건화하는 것이 타이밍 불규칙성과 무관한 측면을 포함하여 모델의 정상 트래픽 표현을 더 정밀하게 만들어, 부차적으로 RTP 페이로드 콘텐츠에 대한 판별력을 높였을 가능성입니다. 둘째, 공격자가 타이밍 보존을 의도했음에도 불구하고 스트림 교체 순간에 미세한 패킷 간 섭동이 발생했을 가능성입니다. 두 가설을 구분하려면 어텐션 가중치 분석이나 타이밍에만 섭동을 주는 통제 실험이 필요하며, 이는 향후 연구 과제로 남겨둡니다.
A notable observation is that time-aware positional encoding (Section~\ref{sec:pe_ablation}) improves RTP detection by AUPRC $+0.043$ despite RTP stream replacement deliberately \emph{preserving} inter-packet timing, implying the gain does not arise from direct detection of timing anomalies. We offer two non-exclusive hypotheses. First, conditioning the Transformer's self-attention on real IPI data may yield a more precise learned representation of normal traffic globally --- including aspects unrelated to timing irregularities --- such that the content stream becomes more discriminative for RTP payload reconstruction as a secondary benefit. Second, despite the attacker's intent to preserve timing, the frame-replacement mechanism may itself introduce subtle inter-arrival perturbations at the moment of stream substitution, which a model conditioned on raw timing data may be more sensitive to than a rank-only sinusoidal baseline. Distinguishing between these hypotheses would require attention-weight analysis or a controlled ablation that perturbs timing while holding payload content fixed; we leave this to future work.

\subsection{Practical Deployment Considerations}
% 제안된 방법은 실제 차량 환경으로의 실무적인 배포를 염두에 두고 설계되었습니다. 시간 인지형 위치 인코딩을 적용하더라도 모델은 약 85만 7천 개의 학습 가능한 파라미터로 구성되어 있으며(Section \ref{sec:ablation}에서 평가된 프로토콜 인지형 RTP 헤더 스트림 제외), 이는 IPI 사영 모듈이 두 개의 작은 선형 레이어로만 구성되어 있기 때문에 기존 사인파형 모델(84만 개) 대비 파라미터 증가량이 1만 7천 개 미만($<$2%)에 불과합니다. 이는 32비트 정밀도 기준 약 3.4,MB의 메모리 점유량에 대응 수치로, 차량용 임베디드 하드웨어의 자원 제약 내에 충분히 들어맞습니다. 또한 모델 학습이 차량 운행 중 수집된 정상 트래픽만으로 수행되므로, 별도의 공격 시뮬레이션이나 수동 어노테이션이 필요하지 않습니다.
The proposed method is designed with practical automotive deployment in mind. With time-aware positional encoding, the model comprises approximately 857K trainable parameters (excluding the protocol-aware RTP header stream evaluated in Section~\ref{sec:ablation}), an increase of fewer than 17K parameters ($<$2\%) over the sinusoidal-PE backbone (840K), since the IPI projection consists of only two small linear layers. This corresponds to a memory footprint of approximately 3.4\,MB in 32-bit precision, remaining well within the constraints of automotive-grade embedded hardware. Training is performed exclusively on normal traffic captured during vehicle operation, requiring no attack simulation or manual annotation.

% 슬라이딩 윈도우 추론을 적용할 때 실무적으로 고려해야 할 점은 탐지 결정을 내리기 전에 $N = 64$개의 연속된 패킷을 버퍼링해야 하므로 지연 시간(latency)이 발생한다는 것입니다. CarDS 데이터셋의 경우, 평가된 네트워크 세그먼트의 평균 패킷 간 격차는 마이크로초에서 밀리초 단위이므로, 정상 트래픽 조건에서 하나의 윈도우가 누적되는 시간은 1초 미만입니다. 이러한 지연 시간은 하드 실시간(hard-real-time) 탐지보다는 준실시간(near-real-time) 탐지로도 충분한 대부분의 차량 침입 탐지 유스케이스에서 수용 가능한 수준입니다. 그럼에도 불구하고, 지연 시간에 민감한 환경을 위해 탐지 성능을 유지하면서 윈도우 크기를 줄이는 것은 여전히 해결해야 할 과제로 남아있으며, 이는 향후 연구 과제로 남겨둡니다.
A practical consideration in sliding-window inference is the latency introduced by the requirement to buffer $N = 64$ consecutive packets before issuing a detection decision. In the CarDS dataset, the average inter-packet interval on the evaluated network segments is on the order of microseconds to milliseconds, resulting in a window accumulation time of well under one second under normal traffic conditions. This latency is acceptable for most automotive intrusion detection use cases, in which near-real-time rather than hard-real-time detection is sufficient. Nevertheless, reducing the window size while maintaining detection performance remains an open direction for latency-sensitive deployments, and is left for future investigation.

% 두 번째 배포 관련 고려사항은 임계값 미세 오차에 대한 민감도입니다. 검증 세트에서 도출된 백분위수 임계값이 배포 트래픽으로 그대로 전이된다는 보장이 없습니다. 차량 노후화, 소프트웨어 업데이트, 또는 학습 데이터에 반영되지 않은 주행 조건으로 인해 재구성 오차 분포가 드리프트될 수 있습니다. Section~\ref{sec:threshold}에서 문서화한 차등 임계값 강건성이 이런 맥락에서 실질적 의미를 가집니다. 사인파형 모델은 최적 임계값에서 단 한 백분위수만 벗어나도 RTP 탐지가 거의 완전히 붕괴되는 반면, 시간 인지형 모델은 최적점($P_{99}$)에서 두 단계 이상 여유를 두고 안정적으로 유지됩니다. 더 넓은 안정 마진을 갖는 것은 배포 환경에서의 드리프트가 탐지 성능을 임계 붕괴 지점으로 밀어낼 위험을 실질적으로 줄여줍니다.
A second deployment-relevant consideration is sensitivity to threshold miscalibration. Validation-set percentile thresholds are not guaranteed to transfer exactly to deployment traffic: the reconstruction error distribution may drift due to vehicle aging, software updates, or driving conditions not represented in the training data. The differential threshold robustness documented in Section~\ref{sec:threshold} is therefore practically significant. The sinusoidal-PE baseline's RTP detection collapses almost entirely with a shift of a single percentile beyond its optimum, whereas the time-aware model's stable region extends substantially beyond its selected operating threshold before meaningful degradation begins. A wider stable margin materially reduces the risk that distribution drift in deployment pushes the threshold past a collapse point.